Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-07-22 analysis/paper/paper.qmd
work on the historical background

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009; @Alexander1998].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001]. Despite some discussions on the policies and operations of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people to the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

@@

# Colonial encounter in Northeastern Taiwan
<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which established a colony in southern Taiwan two years before the Spanish in the north. In 1642, the Dutch defeated the Spanish and occupied their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in Taiwan in 1662 [@Andrade2007]. The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and resources collections among different Indigenous groups. For example, the Basay people, an indigenous group active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@Kang2016]. The colonial domination was mostly established through the conquest over indigenous villages, but the colonial control varied depending on the location of indigenous settlements. For most indigenous villages distant from the European colonial base, the colonial control was relatively less, indirect, and even unsuccessful.
<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->

Kavalan, or the present day Yilan, in northeastern Taiwan is an example of slight European presence for most of the time. Kavalan is about 70 kilometers away from the European northern fortress with mountains in between. Local indigenous groups were Kavalan people. According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, suggesting a continuous indigenous habitation until the 19th century. Kavalan people first experienced European powers in 1632, when they were attacked by Spanish traders [@citation]. Later in 1647, a group of Dutch soldiers attacked several Kavalan villages and asked those defeated villages to pay tributes [@Kang2016]. According to Dutch written record, the VOC had built a trading outpost in Trobiawan, located near the coastal area in Yilan, but there might be not much officials at any given time. It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from colonial powers. The European control over the Kavalan must have been largely indirect and limited, which allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes requested by the VOC, such as deerskins or products [@Kang2016].

Before the arrival of the Europeans in northern Taiwan, there had been a regional exchange network involving Chinese merchants, Japanese traders, and indigenous groups in northern and northeastern Taiwan since the 14th century [@citation]. The local indigenous groups traded

>>>>

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jorden2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources based on commercial interest [@Prakash1997]. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009; @Alexander1998].

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan for a better access to Asian market. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001]. Despite some discussions on the policies and operations of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people to the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

@@

# European encounters in Northeastern Taiwan
<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
In 1626, a fleet from the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and built Fort San Salvador at Heping Dao in Keelung, followed by Fort Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had establish a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Despite the intention to establish a colonial city, the colony of San Salvador remained in its early developmental stages, primarily consisting of a presidio (military outpost), a church, and a population of Spanish officers, soldiers, and missionaries [@Borao2009, pp. 103-134; @Berrocal2018]. The area neighboring Fort San Salvador was inhabited by the Taparris and Quimaurris, local indigenous groups who had resided at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. During the Spanish colonization (1626-1642), the colonial processes may not have involved extensive colonialism or asymmetrical power dynamics, with territorial control likely limited to the areas of fortresses. Written records mention that the Spanish appeared to rely on the local indigenous people for exchanging local goods and food supplies [@Borao2001, SIT:170-171]. This is further supported by archaeological evidence at San Salvador [@Berrocal2018], where only a few European materials were found, alongside more indigenous and Chinese items that might be used by the Europeans, indicating cultural entanglement.

<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
In 1642, the Dutch defeated the Spanish and took over their fortresses, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007]. For example, the VOC governors used various colonial institutions to secure their domination, including landdag ceremonies, taxation, village leasehold systems, selection of representatives, and census investigations [@Kang2016; @Andrade1997; @Andrade2007]. After the Dutch settled northern Taiwan, they expanded their powers further to eastern Taiwan. However, their colonial control varied largely across different regions. For villages distant from the European colonial base, direct control was often challenging and mostly unsuccessful. In these cases, Europeans relied on indigenous intermediaries for colonial activities in remote areas. This cooperative interactions was evident in northern Taiwan, where the Dutch worked closely with the Basay people, an active indigenous group in Keelung, who assisted in coordination and collection of tributes among different indigenous groups in northern and northeastern Taiwan and had also worked with Spanish governors earlier [@Kang2016].

<!-- Kavalan as an example -->
Kavalan, the present day Yilan, in northeastern Taiwan is an example of slight European presence and indirect colonization for most of the time. Kavalan is about 70 kilometers away from the European northern fortress with mountains in between. Local indigenous groups were Kavalan people. According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, suggesting a continuous indigenous habitation until the 19th century. Kavalan people first experienced European powers in 1632, when they were attacked by a Spanish fleed [@citation]. Later in 1647, a group of Dutch soldiers attacked several Kavalan villages and asked those defeated villages to pay tributes [@Kang2016]. According to Dutch written record, the VOC had built a trading outpost in Trobiawan, located near the coastal area in Yilan, but not much information about the number of officials or its operation. It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from European attacks. The European control over the Kavalan must have been largely indirect and limited, which allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes, such as deerskins or products, requested by the VOC, which was even aborted later due to their refusal to pay [@Kang2016].

<!-- previous trade network -->
Before the arrival of the Europeans in northern Taiwan, there had been a regional exchange network involving Chinese merchants, Japanese traders, and indigenous groups in northern and northeastern Taiwan since the 14th century [@citation]. The local indigenous groups traded
